iSEE: Interactive SummarizedExperiment Explorer

### 2018-04-28 manuscript/main.tex
Reorganized functionalities section to remove redundancies, detail.

\section*{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as transcriptomics.
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions, microarray probe sets).\\

Most existing tools for interactive visualization of biological data are designed for specific assays and analyses, e.g., pRoloc for proteomics \citep{gatto2014mass}, shinyMethyl for methylation \citep{fortin2014shinymethyl}, HTSvis for high-throughput screens \citep{scheeder2017htsvis}.
Opportunities for customization are generally limited, making it difficult to re-use the same visualization software for new technologies or experimental designs where different aspects of the data are of interest.
Moreover, standalone tools such as the Loupe Cell Browser from 10X Genomics \citep{zheng2017massively} do not easily integrate into established analysis pipelines such as those based on the R statistical programming language \citep{rcore2008R}.
This complicates any coordinated use of these tools with a reproducible, transparent, and statistically rigorous analysis.

\begin{figure*}[t]
\includegraphics[width=\textwidth]{pics/Fig1.pdf}
\caption{iSEE uses a customisable multi-panel layout (A) that simultaneously displays one or more panels of various types, where each panel type visualizes a different aspect of the data.
New panels of any type can be added (i), and all panels can be removed, reordered or resized (ii).
Panel types are available to visualize sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v).
For these panel types, each sample is represented as a point, which can be coloured according to sample-level metadata or feature-level observations.
Depending on whether the variables of interest are categorical or numeric, panels will automatically switch between scatter plots, violin plots or ``rectangle plots'' (where each combination of categorical levels is represented by a rectangle with area proportional to the frequency of combination).
Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heatmaps (vii), to visualize experimental observations for multiple features; and feature-level metadata plots.
(B) Information can be transmitted between panels according to a user-specified scheme.
Here, the selection of feature $X$ in the row statistics table determines the y-axis of the feature expression plot, and colours the samples in the reduced dimension plot by the expression of $X$.
Selection of points in the reduced dimension plot (dotted blue line) also determines the samples that are shown in the column data (i.e., sample metadata) plot;
further selection of points in the column data plot determines the samples that are shown in the heatmap.
}
\label{fig:iSEE}
\end{figure*}

\section*{Implementation}
Here, we present the iSEE software package for interactive data exploration.
iSEE is implemented in R using the Shiny framework \citep{chang2017shiny} and exploits data architectures from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
The SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where rows and columns represent genomic features and biological samples, respectively.
For instance, individual assays may represent gene expression matrices of raw counts along with different normalization strategies.
In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; those may include experimental metadata as well as any analysis results.
All of these data aspects can be readily examined alongside one another in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).

Notably, the flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
As a consequence, the widespread adoption of the SummarizedExperiment class and its specialized extensions such as the SingleCellExperiment class enables iSEE to offer immediate interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis workflows and methodologies already available in R/Bioconductor packages.
%In particular, we leverage off the state-of-the-art allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).
% KRA: any other child class like SingleCellExperiment worth mentioning? e.g. the BSseq class

% \section*{Description of the iSEE interface} % FM: if required, "turn on the section"?
The layout of the iSEE user interface is built using the shinydashboard package \citep{chang2018shinydashboard}.
The header contains three dropdown menus that provide constant access to diagnostics of the current application state, documentation, and additional information, respectively. % TODO, FM: maybe we can/should cross-reference to the upcoming sections?
The sidebar of iSEE contains a menu enabling the creation of new plots and tables (referred to as ``panels'') in the interface.
Individual color-coded tabs allow the corresponding panel to be re-ordered, adjusted in width and height, or removed from the user interface.
The main component of the iSEE interface displays a combination of active panels controlled by the user's actions in the current session.
Notably, individual panels can optionally be linked to one another, allowing information to be dynamically transmitted between the panels.
In addition to the package documentation and vignette, users can familiarize themselves with the main elements of the user interface by taking an interactive tour (based on the rintrojs package []ganz[]), allowing them to learn the basic usage mechanisms by doing.

\section*{Functionalities}
Currently, six different types of panels can be generated with iSEE. For each panel, three different set of parameters are available in collapsible boxes: namely, Data parameters control the sources of information specific to each type of plot; Visual parameters direct the appearance of the data points (e.g. color, opacity, size); and Selection parameters adjust the behaviour of point selections transmitted between linked panels.

Individual panel types include:

\begin{itemize}
\item Reduced dimension plots -  Display any two dimensions from precomputed reduced dimension matrices (e.g., PCA, t-SNE) available in the reducedDim slot of a SingleCellExperiment object supplied to iSEE. Notably, the capacity of SingleCellExperiment objects to store any number of reduced dimension matrices enables efficient exploration of high-dimensional datasets.
\item Column data plots - Visualize sample metadata stored in the colData slot of SummarizedExperiment object. Conveniently, both categorical and continuous covariates are seamlessly supported, with the geometry of the plot layers automatically switching between scatter plots, violin plots, or bi-categorical grid plots, according to the nature of the data on both axes.
\item Feature expression plots - Visualize expression values from any assay stored in the SummarizedExperiment object for a particular feature (e.g., gene, protein) within individual samples.
This type of plot often adopts a violin or scatter plot geometry, according to the nature of the second covariate.
\item Row statistics tables - Present the contents of the rowData slot for the SummarizedExperiment object.
These tables are particularly helpful when dynamically linked with other panels, both as summary tables of selected genes and sources of selections to visualise or color in target plots.
\item Row data plots - Visualize the information stored in the rowData slot of a SummarizedExperiment object, following an implementation similar to column data plots.
\item Heat maps - Simultaneously visualize assay data and sample metadata, in the form of color-coded expression matrix where rows represent features and columns represent samples.
As such, heat maps may receive both row and column data inputs from appropriately linked panels, to control the subsetting and highlighting of individual features and samples selected in the respective source panels.
\end{itemize}

The color of individual data points (i.e., samples) can be controlled in different ways in each plot. The default is no color scheme, with the same color (default, black) being applied uniformly to all data points.
However, any column from the metadata available in the SummarizedExperiment may be used, as well as any any assay data. The plotting routines automatically adjust the scale to use, based on the chosen variable.

\subsection*{Dynamic linking between panels}
A key feature of iSEE is the ability to dynamically transmit information between panels (Fig.~\ref{fig:iSEE}B).
Users can define and reorganize arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting panels control the inclusion and appearance of the corresponding data points in receiving panels.
This feature facilitates exploration of the relationships between different aspects of the data.
For example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding -- this is achieved by transmitting information from the reduced dimension plot to any number of feature expression plots.

Importantly, the linking paradigm extends to multiple panels, whereby a panel can transmit to multiple receivers, and a receiving panel can transmit its own selection to another plot.
Chains of linked plots allow users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}.
Nevertheless, iSEE augments this functionality with the advantage of integrating selections based on any feature and column metadata present in a SummarizedExperiment object.
Lastly, both tables and plot panels may be used as sources and targets of dynamically linked selections (e.g., selecting a table row to control the coloring of data points in a plot).

\subsection*{Advanced features and customization}

Instances of iSEE applications may be individually pre-configured to alter the startup layout and state of individual panels.
Conveniently, this feature makes it possible to showcase salient features of any data set immediately upon startup, avoiding the need for any unnecessary user input.
In particular, startup settings include the number of panels of each type, their position, width, and height in the interface, individual variables used for each axis and color scale, as well as dynamic links between panels and even brushes and lasso selections may all be configured and saved as an R script that may be run or shared with collaborators for this purpose.
Example of such customization are readily demonstrated by the tours, available at
\url{https://github.com/LTLA/iSEE2018/tree/master/tours}. %% TODO: is that redundant of the later description of tours? %%
A comprehensive list of configurable startup values is available in the ?defaults documentation.

\subsection*{ExperimentColorMap: custom color maps}

iSEE also offers a host of other customizable features to improve data exploration.
To complement the default discrete and continuous color scales borrowed from the ggplot2 \citep{wickham2009ggplot2} and viridis packages \citep{garnier2018viridis}, respectively, iSEE implements the ExperimentColorMap class, that acts as a container capable of accomodating any number of arbitrary color maps.
Color maps are stored as functions that require a single argument, namely the number of colors that they should return. A notable exception are discrete color maps, that may choose to ignore their argument in order to consistently return a constant vector of named colors corresponding to the levels of a known factor.
Color maps stored in ExperimentColorMap objects can be named and retrieved through intuitive accessor functions that follow a naming convention mirroring that of SummarizedExperiment objects (i.e., assayColorMap, colDataColorMap, rowDataColorMap), and thereby facilitates integration within Bioconductor workflows.

Notably, iSEE uses a ExperimentColorMap object to coordinate the color scale applied to every assay, column data, and row data variable across all plots of a session.
The default ExperimentColorMap does not contain any custom color map, and leverages an elegant fallback mechanism to consistently return a discrete or continuous colormap, even for undefined covariates.

Color maps may be overriden at three different levels:

\begin{itemize}
  \item Each individual assay, column data, and row data covariate can be assigned its own distinct color map. Those color maps in the assays, colData, and rowData slots, respectively.
  \item Shared colormaps can be defined for all assays, all column data, and all row data covariates, respectively. These shared color maps come into effect for variables that do not have an individual color map.
  \item Globally shared color maps can be defined for all discrete or continuous covariate that do not have color maps defined at the first two levels.
\end{itemize}

\subsection*{Linked feature annotations}

Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl or Entrez. % do we need to cite them? at least ENSEMBL has updated citable papers

\subsection*{Density-dependent downsampling for speed}

For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.

@@

iSEE also automatically memorizes the exact R code that was used to generate every plot and the existing links, extending previous work by \cite{marini2016interrepro}.
The full code is accessible to users at any time, using the `Extract the R code'' button from the diagnostics dropdown menu in the header.
A modal window displays the formatted code in a shinyAce-based text editor with syntax-highlighting and copy-pasting capabilities, allowing users to easily reproduce the results of any exploratory analysis, simply by integrating the code reported by iSEE into their own scripts.

Additionally, the ``Examine panel chart'' functionality, in the iSEE diagnostics dropdown menu, displays a graphical representation of the existing linked point selections between the visible panels. This can be very useful in sessions that include a large number of panels, and complex 'gating' strategies.
Lastly, the ``Display panel settings'' button from the same dropdown menu returns the required lines of code that underlie the current setup of currently open panels, that may be used in startup scripts to launch an iSEE instance in any prefered layout.

\subsection*{App tours and deployment}

After completion of an analysis, users may deploy instances of iSEE applications on web-servers, and include a bespoke step-by-step ``tour'' of their data set, guiding the audience through an examination of the salient features in the data.

This normally requires the setup of a Shiny Server, % TODO: FM, do we want to provide some basic instructions on how to do it?, % KRA: I don't think so, at least not in the paper. However, documentation and skill will be a limiting factor to encourage users
as well as the preparation of the data set at hand in the form of a SummarizedExperiment - which many workflows available in Bioconductor adopt as main representation.
To include results obtained from any number of dimensionality reduction algorithms, the object must be further coerced to the SingleCellExperiment class, that makes the appropriate slot available.

Tours can be provided to the iSEE function in the form of simple data frames containing a list of Document Object Model (DOM) elements in the Shiny app used to anchor the corresponding associated message ultimately displayed to the user.
Examples of such tours can be found in the inst/extdata folder of the iSEE package, as well as in the GitHub repository of this manuscript, \url{https://github.com/LTLA/iSEE2018/tree/master/tours}.

%\section*{Demonstration data sets and tours}
\section*{Use cases} % FM: used often in other software articles in F1000res
To demonstrate the functionalities and flexibility of the iSEE package, we applied our method to a number of publicly available datasets.
Details are provided in the tour subfolder of the GitHub repository for this manuscript (\url{https://github.com/LTLA/iSEE2018}), and for each dataset a combination of \_app.R and \_tour.txt is provided.
An instance of iSEE running on the respective datasets is provided as a stand-alone webpages, for demonstration purposes without the need for any software installation.

>>>>

\section*{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as genomics.
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).\\

Most existing tools for interactive visualization of biological data are designed for specific assays and analyses, e.g., pRoloc for proteomics \citep{gatto2014mass}, shinyMethyl for methylation \citep{fortin2014shinymethyl}, HTSvis for high-throughput screens \citep{scheeder2017htsvis}.
Opportunities for customization are generally limited, making it difficult to re-use the same visualization software for new technologies or experimental designs where different aspects of the data are of interest.
Moreover, standalone tools such as the Loupe Cell Browser from 10X Genomics \citep{zheng2017massively} do not easily integrate into established analysis pipelines such as those based on the R statistical programming language \citep{rcore2008R}.
This complicates any coordinated use of these tools with a reproducible, transparent, and statistically rigorous analysis.\\

Here, we present the iSEE software package for interactive data exploration.
iSEE is implemented in R using the Shiny framework \citep{chang2017shiny} and exploits data structures from the open-source Bioconductor project \citep{gentleman2004bioconductor}, specifically the SummarizedExperiment class.
iSEE allows users to simultaneously visualize multiple aspects of a given dataset, including experimental data, metadata and analysis results.
Dynamic linking and point selection facilitate the flexible exploration of interactions between different data aspects.
Additional functionalities include code tracking, intelligent downsampling of large datasets, custom colour scale specification and tour construction.
We demonstrate the capabilities of iSEE by applying it to a diverse range of real datasets.

\section*{Taking a SummarizedExperiment input}
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
Each instance of the SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where rows and columns represent genomic features and biological samples, respectively.
For instance, individual assays may represent gene expression matrices of raw counts along with different normalization strategies.
In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; those may include experimental metadata as well as any analysis results.\\

The flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem.
SummarizedExperiment objects are currently used in analysis pipelines for RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C data \citep{lun2016infrastructure}, amongst others.
Package developers can also easily use the base SummarizedExperiment class to derive new bespoke classes for particular applications, such as the SingleCellExperiment class for single-cell 'omics data.
By accepting SummarizedExperiment objects as input, iSEE immediately offers interactive visualization for a variety of data modalities.
This complements the state-of-the-art analysis workflows and methodologies already available in R/Bioconductor packages. \\

\section*{Interface implementation}

\subsection*{Using a multi-panel layout}
All data aspects stored in a SummarizedExperiment can be readily examined alongside one another in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
The interface layout is built using the shinydashboard package \citep{chang2018shinydashboard}, using colour-coded panels to visualize each data aspect.
Individual panel types include:

\begin{itemize}
\item Column data plots, for visualizing sample metadata stored in the colData slot of SummarizedExperiment object. 
\item Feature expression plots, for visualizing expression values for a particular feature (e.g., gene) across samples from any assay in the SummarizedExperiment object.
Despite its name, this plot type is applicable to data modalities beyond gene expression.
\item Row statistics tables, to present the contents of the rowData slot for the SummarizedExperiment object.
\item Row data plots, for visualizing sample metadata stored in the rowData slot of a SummarizedExperiment object.
\item Heat maps, to visualize assay data for multiple features where samples are ordered by one or more colData fields.
\item Reduced dimension plots, which display any two dimensions from pre-computed dimensionality reduction results (e.g., from PCA or $t$-SNE).
These results are taken from the reducedDim slot of any SingleCellExperiment object supplied to iSEE. 
\end{itemize}

\begin{figure*}[t]
\includegraphics[width=\textwidth]{pics/Fig1.pdf}
\caption{iSEE uses a customisable multi-panel layout (A) that simultaneously displays one or more panels of various types, where each panel type visualizes a different aspect of the data.
Panels of each type are colour-coded for ease of use.
New panels of any type can be added (i), and all panels can be removed, reordered or resized (ii).
Panel types are available to visualize sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v).
Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heatmaps (vii), to visualize experimental observations for multiple features; and feature-level metadata plots.
(B) Information can be transmitted between panels according to a user-specified scheme.
Here, the selection of feature $X$ in the row statistics table determines the y-axis of the feature expression plot, and colours the samples in the reduced dimension plot by the expression of $X$.
Selection of points in the reduced dimension plot (dotted blue line) also determines the samples that are shown in the column data (i.e., sample metadata) plot;
further selection of points in the column data plot determines the samples that are shown in the heatmap.
}
\label{fig:iSEE}
\end{figure*}

Each sample is represented as a point in column data, feature expression and reduced dimension plots.
Similarly, each feature is represented by a point in row data plots.
For these panel types, a scatter plot is automatically produced if the selected variables on the x- and y-axes are both continuous.
If exactly one variable is categorical, points are grouped by the categorical levels and a (horizontal) violin plot is produced with points scattered within each violin.
If both variables are categorical, a ``rectangle plot'' is produced where each combination of categorical levels is represented by a rectangle with area proportional to the frequency of that combination.
Points are scattered randomly within each rectangle.
For ease of interpretation, the rectangle plot collapses to a mirrored bar plot when one of the categorical variables only has one level.

\subsection*{Custom panel colouring}
Sample-based points can be coloured according to the values of any sample-level metadata field in the colData or by the assay values of a selected feature.
Similarly, feature-based points can be coloured according to any feature-level metadata field in the rowData.
Heatmaps are coloured according to the expression values of the selected features in the chosen assay,
with additional colouring of each of the colData fields used to order the samples.
In all cases, the variable to use for colouring can be dynamically selected for each plot.
This enables users to easily examine relationships between different variables in a single plot. \\

By default, colour maps for categorical and continuous variables are taken from the ggplot2 \citep{wickham2009ggplot2} and viridis packages \citep{garnier2018viridis}, respectively.
However, iSEE also implements the ExperimentColorMap class, which allows users to specify arbitrary colour maps for particular variables.
Each colour map is a function that returns a vector of distinct colours of a specified length, and will be called whenever the associated variable is used for point colouring in a particular panel.
The returned colours will be mapped to factor levels for categorical variables, or used in colour interpolation for continuous variables.
For categorical variables, the function may also return a constant vector of named colors corresponding to the levels of a known factor.
Colour maps can be specified for individual variables; for all assays, all column data variables, or all row data variables (with different functions for continuous or categorical variables); or for all categorical or continuous variables.
This provides a convenient yet flexible mechanism for customization of colouring schemes within the interface.

\subsection*{Dynamic linking between panels}
A key feature of iSEE is the ability to dynamically transmit information between panels (Fig.~\ref{fig:iSEE}B).
Users can define and reorganize arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting panels control the inclusion and appearance of the corresponding data points in receiving panels.
This feature facilitates exploration of the relationships between different aspects of the data.
For example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding -- this is achieved by selecting points in a reduced dimension plot (using the standard rectangular brush or a lasso) and transmitting that selection to any number of feature expression plots.\\

This linking paradigm extends to multiple panels, whereby a panel can transmit to multiple receivers, and a receiving panel can transmit its own selection to another plot.
Chains of linked plots allow users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}.
With iSEE, this concept is extended to any assay data, feature-level or sample-level metadata present in a SummarizedExperiment object, providing a powerful framework for interrogating multiple interactions between data aspects.
Row statistics tables can also transmit to various plot types, by selecting a table row to control the coloring of sample-based points;
or by defining a subset of features to visualize in a heatmap.
Furthermore, row data plots can transmit to row statistics tables, whereby selection of points in the former will subset the latter.

@@

iSEE automatically memorizes the exact R code that was used to generate every plot, extending previous work by \cite{marini2016interrepro}.
This code is fully accessible to users at any time during the run-time of the interface.
By integrating the code reported by iSEE into their own scripts, users to easily reproduce the results of any exploratory analysis.
Similarly, the code required to reproduce the current state of the interface can also be generated.
This can be used in startup scripts to launch an iSEE instance in any preferred layout, including the panel organization, variable selection, colouring schemes, links between panels and even individual brushes and lasso selections.

\subsection*{Additional functionalities}
Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl or Entrez. % do we need to cite them? at least ENSEMBL has updated citable papers
For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
Users can also include a bespoke step-by-step ``tour'' of their data set via the rintrojs package \cite{ganz}, guiding the audience through an examination of the salient features in the data.

\section*{Use cases} % FM: used often in other software articles in F1000res
To demonstrate the functionalities and flexibility of the iSEE package, we applied our method to a number of publicly available datasets.
Details are provided in the tour subfolder of the GitHub repository for this manuscript (\url{https://github.com/LTLA/iSEE2018}), and for each dataset a combination of \_app.R and \_tour.txt is provided.
An instance of iSEE running on the respective datasets is provided as a stand-alone webpages, for demonstration purposes without the need for any software installation.
